---
title: "An Brief Introduction to `MBalance`"
author: "Yimin Dai"
date: "`r Sys.Date()`"
bibliography: MBalance.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Brief Introduction to `MBalance`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `MBalance` implements the Mahalanobis balancing method for estimating average treatment effect (ATE) and average treatment effect on the treated group or control group (ATT or ATC) on observational data. Our method has good performance (low bias, small balancing diagnostics) in "Bad Overlap" situations (e.g., $E\{Pr(T = 1 \mid X)\}$ closes to 0 or 1, $E\{ X \mid T = 1\}$ and $E\{ X \mid T = 0\}$ are significantly different, high-dimensional covariate).

```{r}
# Install the "Mbalance" package from github.
# if (!require(c("devtools","WeightIt","MASS","cobalt"))){
#     install.packages(c("devtools","WeightIt","MASS","cobalt"))
# }
# devtools::install_github("yimindai0521/Mbalance")
library(MBalance)
library(WeightIt)
library(MASS)
library(sbw)
library(cobalt)
```

# Implementation of other covariate balancing methods

We provide function `covbal`,`sbwbal` to implement Mahalanobis balancing method and other covariate balancing methods, including inverse probability weighting (ps) in @robins1994estimation, entropy balancing (ebal) in @hainmueller2012entropy, covariate balancing propensity score (cbps) in  @imai2014covariate, energy balancing (energy) in @huling2020energy, minimal dispersion approximate balancing weights (sbw) in @wang2020minimal. 

```{r}
covbal <- function(X, Tr, Y){
  data.matrix <- data.frame(X , factor(Tr))
  sample.size <- dim(X)[1]
  dimension   <- dim(X)[2]
  
  character   <- names(data.matrix)
  for(j in 1:(dimension+1)){character[j] <- paste(character[j])}
  myformula   <- as.formula(paste(character[1 + dimension],paste(" ~ ", paste(character[1:dimension], collapse= "+"))))
  
  ps.weight   <- weightit(myformula , data = data.matrix, estimand = "ATE", method = "ps")$weights
  ps.ate      <- t(ps.weight[Tr == 1] / sum(ps.weight[Tr == 1])) %*% Y[Tr == 1] - t(ps.weight[Tr == 0] / sum(ps.weight[Tr == 0])) %*% Y[Tr == 0]
  
  ebal.weight <- weightit(myformula , data = data.matrix, estimand = "ATE", method = "ebal")$weights
  ebal.ate    <- t(ebal.weight[Tr == 1] / sum(ebal.weight[Tr == 1])) %*% Y[Tr == 1] - t(ebal.weight[Tr == 0] / sum(ebal.weight[Tr == 0])) %*% Y[Tr == 0]
  ebal.weight[ebal.weight == 0] <- 1
  
  cbps.weight <- weightit(myformula , data = data.matrix, estimand = "ATE", method = "cbps")$weights
  cbps.ate    <- t(cbps.weight[Tr == 1] / sum(cbps.weight[Tr == 1])) %*% Y[Tr == 1] - t(cbps.weight[Tr == 0] / sum(cbps.weight[Tr == 0])) %*% Y[Tr == 0]
  
  energy.weight <- weightit(myformula , data = data.matrix, estimand = "ATE", method = "energy")$weights
  energy.ate    <- t(energy.weight[Tr == 1] / sum(energy.weight[Tr == 1])) %*% Y[Tr == 1] - t(energy.weight[Tr == 0] / sum(energy.weight[Tr == 0])) %*% Y[Tr == 0]

  MB.weight   <- rep(NA,sample.size)
  MB1.result  <- MB(x = X, treat = Tr, group1 = 0, outcome = rep(0,sample.size), method = "MB")
  MB2.result  <- MB(x = X, treat = Tr, group1 = 1, outcome = rep(0,sample.size), method = "MB")
  MB.weight[Tr == 0] <- sum(Tr == 0) * MB1.result$weight
  MB.weight[Tr == 1] <- sum(Tr == 1) * MB2.result$weight
  MB.ate      <- t(MB.weight[Tr == 1] / sum(MB.weight[Tr == 1])) %*% Y[Tr == 1] - t(MB.weight[Tr == 0] / sum(MB.weight[Tr == 0])) %*% Y[Tr == 0]

  
  return(list(weight = list(ps = ps.weight, ebal = ebal.weight, cbps = cbps.weight, energy = energy.weight, MB = MB.weight), ate = c(ps = ps.ate, ebal = ebal.ate, cbps = cbps.ate, energy = energy.ate, MB = MB.ate)))
}

sbwbal <- function(X, Tr, Y){
  data.matrix <- data.frame(X , factor(Tr), Y)
  dimension   <- dim(X)[2]
  bal = list()
  bal$bal_gri = c(1e-04, 0.001, 0.002, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1)
  character   <- names(data.matrix)
  bal$bal_cov <- character[1:dimension]
  sbw.result <- FALSE
  while(sum(dim(as.matrix(sbw.result))) != 15){
    sbw.result  <- tryCatch(sbw.result <- sbw(dat = data.matrix, ind = character[1 + dimension], bal = bal, out = character[2 + dimension], par = list(par_est = "ate")), error = function(e) { skip_to_next <<- FALSE})
    bal$bal_gri <- bal$bal_gri[-1]
  }
  sbw.weight <- sbw.result$dat_weights$sbw_weights
  sbw.ate    <- t(sbw.weight[Tr == 1] / sum(sbw.weight[Tr == 1])) %*% Y[Tr == 1] - t(sbw.weight[Tr == 0] / sum(sbw.weight[Tr == 0])) %*% Y[Tr == 0]
   return(list(weight = sbw.weight, ate = sbw.ate))
}
```

# Generating data 

We consider three different simulation settings. The first simulation will be Scenario C in our paper. We first generate treatment indicator $T$ from $Bernoulli(0.5)$ for each observation. The observed covariates depend on treatment assignment. If $T = 1$, then $X \sim N(1, \Sigma)$ where the jth row and kth column of $\Sigma$ is $\rho^{|j−k|}$ and we set $\rho = 1/2$, otherwise, $X \sim N(0,I)$. The observed outcome is generated from: $Y(T) = (1 - T)(X_1 + ... + X_6) + T(X_1 + ... + X_6) / 2$. In this setting, $E\{ X \mid T = 1\}$ and $E\{ X \mid T = 0\}$ are significantly different and thus it's a "Bad Overlap" situation.


```{r}
si.data.bad <- function(dimension = 10, sample.size = 200){
  covmatrix   <- matrix(0 , dimension , dimension)
  for(i in 1:dimension){for(j in 1:dimension){covmatrix[i,j] <-2^{-abs(i!=j)}}}
  treat  <- rbinom(sample.size,1,0.5)
  z1     <- mvrnorm(sample.size, mu = rep(1,dimension), Sigma = covmatrix)
  z0     <- mvrnorm(sample.size, mu = rep(0,dimension), Sigma =         diag(dimension))
  X      <- treat * z1 + (1-treat) * z0
  noise  <- runif(sample.size)
  Y      <- (1 + treat) * apply(X[,1:10],1,sum) + noise
  return(list(X = X, treat = treat, Y = Y))
}
```

The second simulation setting corresponds Scenario D to our paper. The observed covariates are simulated by $X \sim N(1,I)$. The treatment indicator is simulated from $T ∼ Bernoulli(\pi(X))$ with $π(X) = 1/(1+19 \exp(X_1+···+X_{10}−10))$. The outcome is simulated from $Y = (1+T)(X_1+···+X_{10}) + \epsilon$, where $\epsilon \sim N(0,1)$. In this setting, $E\{Pr(T = 1 \mid X)\}$ closes to 0 or 1 and thus it's a "Bad Overlap" situation.

```{r}
si.data.badover <- function(dimension = 10, sample.size = 1000){
  X      <- matrix(rnorm(sample.size * dimension) + 1, nrow = sample.size, ncol = dimension)
  ps     <- 1 / (1 + 19 * exp( apply(X[,1:10],1,sum) - 10) )
  treat  <- rep(NA,sample.size)
  for(i in 1:sample.size){treat[i]  <- rbinom(1,1,ps[i])}
  noise   <- runif(sample.size)
  Y       <- (1 + treat) * apply(X[,1:10],1,sum) + noise
  return(list(X = X, treat = treat, Y = Y))
}
```

We finally study the performance of our method in high-dimensional data. The final simulation corresponds to Scenario E in our paper. We first generate the observed data from $X ∼ N(0,\Sigma)$ where the jth row and kth column of $\Sigma$ is $\rho^{|j−k|}$ and we set $\rho = 1/2$. The treatment index is generated by $T ∼ Bernoulli(π(X))$ with $\pi(X) = 1/(1 + \exp(X_1 + (X_2 + ... + X_6) / 2))$. The outcome is generated from: $Y(T) = (1 - T)(X_1 + ... + X_6) + T(X_1 + ... + X_6) / 2$.

```{r}
si.data.hd <- function(dimension = 100, sample.size = 200){
  covmatrix   <- matrix(0 , dimension , dimension)
  for(i in 1:dimension){for(j in 1:dimension){covmatrix[i,j] <-2^{-abs(i-j)}}}
  X      <- mvrnorm(sample.size, mu = rep(1,dimension), Sigma = covmatrix)
  ps     <- 1 / (1 + exp( apply(X[,1:6],1,sum) / 2) )
  treat  <- rep(NA,sample.size)
  for(i in 1:sample.size){treat[i]  <- rbinom(1,1,ps[i])}
  noise  <- runif(sample.size)
  Y      <- (1 + treat) / 2 * apply(X[,1:6],1,sum) + noise
  return(list(X = X, treat = treat, Y = Y))
}
```

# `MB`: estimating average treatment effect

We will use function \texttt{love.plot} in R package @Greifer2022cobalt to plot ASMD to compare the performance for different covariate balancing methods.

```{r}
##generating data from Scenario C
set.seed(1999)
data   <- si.data.bad()
result <- covbal(data$X,data$treat,data$Y)
sbw.re <- sbwbal(data$X,data$treat,data$Y)

##plot ASMD
dimension   <- dim(data$X)[2]
data.matrix <- data.frame(data$X , factor(data$treat))
character   <- names(data.matrix)
for(j in 1:(dimension+1)){character[j] <- paste(character[j])}
myformula   <- as.formula(paste(character[1 + dimension],paste(" ~ ", paste(character[1:dimension], collapse= "+"))))
love.plot(myformula, data = data.matrix, estimand = "ATE",
                       stats = c("mean.diffs"),
                       weights = list(w1 = result$weight$MB,
                                      w2 = result$weight$ps,
                                      w3 = result$weight$ebal,
                                      w4 = result$weight$cbps,
                                      w5 = result$weight$energy,
                                      w6 = sbw.re$weight),
                       var.order = "unadjusted",
                       abs = TRUE,
                       line = TRUE, 
                       thresholds = c(m = .1),
                       sample.names = c("Unweighted", "MB", "PS", "ebal", "cbps", "energy","sbw"),
                       limits = list(m = c(0,1.5)),
                       wrap = 20,
                       position = "top")

##generating data from Scenario D
data   <- si.data.badover()
result <- covbal(data$X,data$treat,data$Y)
sbw.re <- sbwbal(data$X,data$treat,data$Y)

##plot ASMD
dimension   <- dim(data$X)[2]
data.matrix <- data.frame(data$X , factor(data$treat))
character   <- names(data.matrix)
for(j in 1:(dimension+1)){character[j] <- paste(character[j])}
myformula   <- as.formula(paste(character[1 + dimension],paste(" ~ ", paste(character[1:dimension], collapse= "+"))))
love.plot(myformula, data = data.matrix, estimand = "ATE",
                       stats = c("mean.diffs"),
                       weights = list(w1 = result$weight$MB,
                                      w2 = result$weight$ps,
                                      w3 = result$weight$ebal,
                                      w4 = result$weight$cbps,
                                      w5 = result$weight$energy,
                                      w6 = sbw.re$weight),
                       var.order = "unadjusted",
                       abs = TRUE,
                       line = TRUE, 
                       thresholds = c(m = .1),
                       sample.names = c("Unweighted", "MB", "PS", "ebal", "cbps", "energy","sbw"),
                       limits = list(m = c(0,1)),
                       wrap = 20,
                       position = "top")
```

We can find that ASMD of MB is significantly lower than other covariate balancing methods.

We repeat 100 times to achieve bias for each method.
```{r}
mainf1 <- function(iteration = 100){
  result1 <- matrix(0,iteration,5)
  result2 <- matrix(0,iteration,5)
  for(i in 1:iteration){
    data1       <- si.data.bad()
    data2       <- si.data.badover()
    result1[i,] <- covbal(data1$X,data1$treat,data1$Y)$ate
    result2[i,] <- covbal(data2$X,data2$treat,data2$Y)$ate
    }
  return(list(bad = list(ps_bias = mean(result1[,1]) - 5, ebal_bias = mean(result1[,2]) - 5, cbps_bias = mean(result1[,3]) - 5, energy_bias = mean(result1[,4]) - 5, MB_bias = mean(result1[,5]) - 5), badover = list(ps_bias = mean(result2[,1]) - 10, ebal_bias = mean(result2[,2]) - 10, cbps_bias = mean(result2[,3]) - 10, energy_bias = mean(result2[,4]) - 10, MB_bias = mean(result2[,5]) - 10)
              ))
}

set.seed(0521)
options(warn = -1)
mainf1(100)

```

We can find that our method has lower bias than other methods because our purposed method has significantly lower imbalance and thus greatly outperforms other methods.

#`hdMB`: estimating average treatment effect in high-dimensional data.

```{r}
set.seed(0521)
data <- si.data.hd()
dimension <- dim(data$X)[2]
result1 <- hdMB(x = data$X, treat = data$treat, outcome = data$Y, GASMD.bound = 0.2, group1 = 1)

plot(result1$GMIM)
threshold   <- 4
index       <- rank(result1$GASMD) >= (dimension - threshold + 1)
result1     <- MB(x = data$X[,index], treat = data$treat, outcome = data$Y, group1 = 1)


result0 <- hdMB(x = data$X, treat = data$treat, outcome = data$Y, GASMD.bound = 0.05, group1 = 0)
plot(result0$GMIM)
threshold   <- 16
index       <- rank(result0$GASMD) >= (dimension - threshold + 1)
result0     <- MB(x = data$X[,index], treat = data$treat, outcome = data$Y, group1 = 1)

##An estimate of ATE
result1$AT - result0$AT
```

# References







